A tool set for random number generation on GPUs in R

### 2021-12-30 pkg/gpuRandom/inst/documents/paper1_2021_5/paper1.Rnw
a few edits
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/diseasemapping@3079 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

%The function \fct{clrngMrg31k3pCreateStreams} from \proglang{clRNG} creates streams on the host using the MRG31k3p RNG. If we want to use the streams in work-items on a GPU device, the streams have to be copied from the host to the global memory of the GPU device, and then each work-item copies the current states only of the streams to its private memory. 
\pkg{clrng} is able to create streams both on the host and on the GPU device. The \fct{createStreamsCpu} function does the former. The \proglang{R} output below creates 4 streams on the host. 
%is an interface function of \fct{clrngMrg31k3pCreateStreams} that creates streams on host as \proglang{R} matrices, see the \proglang{R} output below, where we created 4 streams on the host using \fct{createStreamsCpu}, after transposing \code{myStreamsCpu}, each column of the matrix is a stream, creating streams on host makes it easy for us to view and check the streams behavior. Substreams are not created in \pkg{clrng} because currently we use one stream per work-item. 
<<createStreams_CPU, message=FALSE>>=
# creating streams on CPU
library("clrng")
myStreamsCpu <- createStreamsCpu(n=4, initial=12345)
t(myStreamsCpu)
@
\fct{createStreamsCpu} has two arguments:
\begin{itemize}
\itemsep0em 
  \item \code{n}: number of streams to create.
  \item \code{initial}: a vector of six 31-bit integers, recycled if length is shorter. Default is set to \code{c(12345, 12345, 12345, 12345, 12345, 12345)}.
\end{itemize}
We move streams to the GPU by converting them to a `\code{vclMatrix}'. \fct{createStreamsGpu} creates streams directly on the GPU and returns a `\code{vclMatrix}', which makes it slightly more efficient when the number of streams is large. Users may choose to use either \fct{createStreamsCpu} or \fct{createStreamsGpu} and stick with that choice for their entire program, as we want to avoid the very small possibility of overlap between the streams with different initial seeds. The `\code{initial}' argument should be set only once at the very beginning of the program in each \proglang{R} session if users want to supply their own initial seed of the stream creator, or never, in which case users would go with the package default initial seed. After the `\code{initial}' argument is set, we can ignore the `\code{initial}' argument and call \fct{createStreamsCpu} or \fct{createStreamsGpu} once or several times to create various matrices of streams in the same program.

%However, any data is lost when the \proglang{R} session is terminated. %\fct{createStreamsGpu} is adapted from the clrngMrg31k3pCreateStreams() and creates streams directly on a GPU device. Setting a seed same as the MRG31k3p default seed in \fct{createStreamsGpu}, we produce on device exactly the same streams as those created on host as follows. Notice that the code is just for showing \fct{createStreamsGpu} and \fct{createStreamsCpu} are \proglang{R} interface functions for \fct{clrngMrg31k3pCreateStreams} on GPU and CPU respectively, in practice it is not recommended to create a new stream that has the same initial seed as another existing stream in one simulation program.
<<createStreams_Gpu, eval=FALSE>>=
# creating streams on GPU
myStreamsGpu = vclMatrix(myStreamsCpu)
@
<<createStreams_Gpu2>>=
myStreamsGpu2 = createStreamsGpu(n=4, initial=12345)
@

@@

\subsection{Using streams to generate uniform random numbers}
The streams created sequentially are then used by the GPU's work-items to generate random numbers. In \pkg{clrng} each work-item takes one distinct stream to generate random numbers, so the number of streams should always be at least the number of total work-items in use. The main part of the kernel (\proglang{OpenCL} functions for execution on the device) for generating uniform random numbers is shown in Listing \ref{lst:uniformkernel}. Users can set \code{verbose=2} in \fct{runif} to print out the kernels. Kernels are written in the \proglang{OpenCL C} language, in which \code{__kernel} declares a function as a kernel, pointer kernel arguments must be declared with an address space qualifier, for example \code{__global} and \code{__local}. Here the pointers to \code{streams} and to the output matrix \code{out} on the global memory are passed to the kernel as arguments. \code{Nrow} and \code{Ncol} represent row and column number of the matrix \code{out} respectively. \code{Npad} is the internal number of columns of \code{out}, \code{NpadStreams} is the internal number of columns of \code{streams} (these variables are defined in macros not shown here). \code{index} represents the position of a work-item when mapping it to a one-dimensional data structure. Each stream's current state is copied to the private memory of each work-item by the function \fct{streamsToPrivate}, in which \code{g1} and \code{g2} point to the first three and second three elements of stream states, respectively, and \code{startvalue} tells the starting position of the particular stream to be copied to its corresponding work-item. The function \fct{clrngMrg31k3pNextState} generates an uniform random integer named \code{temp} between 1 and 2147483647,  the value of \code{temp} is then scaled to be in the interval $(0,1)$ by multiplying it by a constant \code{mrg31k3p\_NORM\_cl}, which is defined to be $1/2147483648$ depending on the precision type. If to generate random integers, \code{temp} is not scaled. At the end of generating random numbers, streams are transferred back to global memory through the function \fct{streamsFromPrivate}.

@@

int Drow, Dcol;
uint g1[3], g2[3];
float temp, fact = mrg31k3p_NORM_cl;

@@

>>>>

%The function \fct{clrngMrg31k3pCreateStreams} from \proglang{clRNG} creates streams on the host using the MRG31k3p RNG. If we want to use the streams in work-items on a GPU device, the streams have to be copied from the host to the global memory of the GPU device, and then each work-item copies the current states only of the streams to its private memory. 
\pkg{clrng} is able to create streams both on the host and on the GPU device. The \fct{createStreamsCpu} function does the former. The \proglang{R} output below creates 4 streams on the host. 
%is an interface function of \fct{clrngMrg31k3pCreateStreams} that creates streams on host as \proglang{R} matrices, see the \proglang{R} output below, where we created 4 streams on the host using \fct{createStreamsCpu}, after transposing \code{myStreamsCpu}, each column of the matrix is a stream, creating streams on host makes it easy for us to view and check the streams behavior. Substreams are not created in \pkg{clrng} because currently we use one stream per work-item. 
<<createStreams_CPU, message=FALSE>>=
# creating streams on CPU
library("clrng")
myStreamsCpu <- createStreamsCpu(n=4, initial=rep(12345, 6))
t(myStreamsCpu)
@
\fct{createStreamsCpu} has two arguments:
\begin{itemize}
\itemsep0em 
  \item \code{n}: number of streams to create.
  \item \code{initial}: a vector of six 31-bit integers, recycled if length is shorter. Default is set to \code{c(12345, 12345, 12345, 12345, 12345, 12345)}.
\end{itemize}
We move streams to the GPU by converting them to a `\code{vclMatrix}'.
<<createStreams_Gpu, eval=FALSE>>=
myStreamsGpu = vclMatrix(myStreamsCpu)
@
Equivalently, 
 \fct{createStreamsGpu} creates streams directly on the GPU and returns a `\code{vclMatrix}', which makes it slightly more efficient when the number of streams is large.
<<createStreams_Gpu2>>=
myStreamsGpu2 = createStreamsGpu(n=4, initial=rep(12345, 6)))
@

Users should run either \fct{createStreamsCpu} or \fct{createStreamsGpu} at the beginning of a task and retain these streams for all of the random number generation the task requires.  Creating new streams with different initial states introduces the very small possibility of overlap between the original and new streams. If an \proglang{R} session is restarted during a task, perhaps because a preliminary analysis is done following an initial set of simulations and the full set of simulations continued at a later date, the state of the streams should be transferred to the CPU and saved to disk.  The `\code{initial}' argument to the stream creator should be used with caution and we recommend using the default.  It bears repeating that two sets of streams with different initial states may produce overlapping numbers, the probability is small but increases when the number of streams is large.

%However, any data is lost when the \proglang{R} session is terminated. %\fct{createStreamsGpu} is adapted from the clrngMrg31k3pCreateStreams() and creates streams directly on a GPU device. Setting a seed same as the MRG31k3p default seed in \fct{createStreamsGpu}, we produce on device exactly the same streams as those created on host as follows. Notice that the code is just for showing \fct{createStreamsGpu} and \fct{createStreamsCpu} are \proglang{R} interface functions for \fct{clrngMrg31k3pCreateStreams} on GPU and CPU respectively, in practice it is not recommended to create a new stream that has the same initial seed as another existing stream in one simulation program.

@@

\subsection{Generating uniform random numbers}

The streams created sequentially are used by work-items on a GPU (the GPU analogy of CPU cores) to generate random numbers. In \pkg{clrng} each work-item takes one distinct stream to generate random numbers, so the number of streams should always be at least the number of total work-items in use. The main part of the kernel (functions for execution on the GPU) for generating uniform random numbers is shown in Listing \ref{lst:uniformkernel}. Kernels are written in the \proglang{C}-like \proglang{OpenCL} language, in which \code{__kernel} declares a function as a kernel, and the _\code{__global} prefix to the pointer kernel arguments specifies that they point to global memory space accessible to all work items.  Users can set the argument \code{verbose=2} in the random number generator to print out the kernel, which is slightly more complex than the code in Listing \ref{lst:uniformkernel}.

Here the pointers \code{streams} and \code{out} refer to the streams and the  output matrix respectively, which are stored in global memory. \code{Nrow} and \code{Ncol} represent row and column number of the matrix \code{out} respectively.   \code{Npad} is the "internal" number of columns, and \code{out} is an \code{Nrow} by \code{Ncol} submatrix of a larger \code{Nrow} by \code{Npad} matrix.

Each stream's current state is copied to the private memory of each work-item by the function \fct{streamsToPrivate}, in which \code{g1} and \code{g2} point to the first three and second three elements of stream states.
The object \code{startvalue} gives the position of the work item's stream in the \code{streams} object and is computed with the
\code{NpadStreams} object defined in a macro not shown here.

The function \fct{clrngMrg31k3pNextState} generates an uniform random integer between 1 and 2147483647,  and then scaled to be in the interval $(0,1)$ by multiplying it by a constant \code{mrg31k3p\_NORM\_cl} (defined to be $1/2147483648$). If to generate random integers, \code{temp} is not scaled. At the end of generating random numbers, streams are transferred back to global memory through the function \fct{streamsFromPrivate}.

@@

int Drow, Dcol;
uint g1[3], g2[3];
double temp, fact = mrg31k3p_NORM_cl;
const int startvalue = (get_global_id(0) + get_global_size(0) * get_global_id(1)) * NpadStreams;

@@

for(Drow = get_global_id(0); Drow < Nrow;
    Drow += get_global_size(0)){

    for(Dcol = get_global_id(1); Dcol < Ncol; 
        Dcol += get_global_size(1)){
